---
title: "Data Jujutsu II -- PhD Trends"
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
author:
  name: Stefano Allesina and Graham Smith
  affiliation: University of Chicago
date: August 26, 2021
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
chapter: 3--J2
thanks: "This document is included as part of the Advanced Computing tutorial packet for the BSD qBio Bootcamp, University of Chicago, 2021. **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: sallesina@uchicago.edu."
---

```{r setup, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

## Description of the data

Every year, the National Science Foundation sponsors a very large survey (with almost complete sampling) of the PhD graduates, the *Survey of Earned Doctorates* (SED). They publish statistics on the number of PhD, and report PhD completion by gender, field, ethnic background, etc. In particular, table 16 reports the number of PhD awarded in total and divided by sex for each field of study. We are going to attempt reading the table directly from the `xlsx` files that are published by NSF.

## The challenge

#### 1. Read the data

The file `urls_and_skip_NSF_SED.csv` reports the location (`url`) of the excel files for the years 2013-1018, as well as the number of lines to skip (`skip`) and the number of lines to read (`read`) for best results. Read the documentation of `read_xlsx` from the library `readxl` to see how to read the file while skipping a few lines and capping the total number of lines to be read. 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
library(tidyverse)
library(readxl)
read_csv("urls_and_skip_NSF_SED.csv")
```

Read all the files, building the tibble `sed` retaining only the field and the total for each year:
  
```{r, eval=TRUE, message=FALSE, warning=FALSE, echo=TRUE}
source("solution_PhD_trends.R") # this is the code you have to write!
sed
```

#### 2. Standardize names and filter

Notice that there are very many fields, and that the names of some fields have changed through the years (e.g., `Neurosciences, neurobiology`, `Neurosciences and neurobiology`). The file `lookup_fields_filter.csv` contains two columns: retain all the records for the fields specified in the table, and use the column `name_to_use` to standardize the names of the fields.  You should end up with 18 fields (all well-represented at U of C) as well as the data for `All fields` and `Biomedical sciences`.
  
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
sed <- sed %>% inner_join(read_csv("lookup_fields_filter.csv"))
```

```{r, eval=TRUE, echo=TRUE}
sed
```

#### 3. Plot the time series

Write a generic function for plotting the number of PhDs awarded for all the fields in a tibble. For example, here are the trends for all PhDs and all PhDs in biomedical sciences:

```{r, echo=TRUE, eval = TRUE}
sed %>% 
  filter(name_to_use %in% c("All fields", "Biomedical sciences")) %>% 
  plot_PhD_in_time()
```
  
#### 4. Fields that have changed the most

Some fields have grown considerably in the past 6 years, while some have shrunk. For example:

```{r, echo=TRUE, eval = TRUE}
sed %>% 
  filter(name_to_use %in% c("Cell biology", "Biostatistics")) %>% 
  plot_PhD_in_time()
```

  Find the fields for which the ratio between the maximum number of PhDs and the minimum number of PhDs for the period considered is the largest.

#### 5. Correlation between time series [Optional]

Compute the correlation between the time series of any two fields. Which fields have changed in synchrony? Plot the matrix of correlations using `geom_tile`.

#### 6. Order the matrix [Optional, requires some math] 

Find a good ordering for the matrix by plotting the field according the the eigenvector of the correlation matrix corresponding to the largest eigenvalue.

## Hints & Nifty tricks

* If you don't want to store the downloaded zip file, use a temporary file (it will be deleted by `R` automatically once you call `unlink()`)

* Some lines are empty: use something like `filter(!is.na(field))` to get rid of them. 

* For each year, you only need to store the number of PhD awarded.

